---
title: "Human chromosome biased"
author: "Laura Biggins"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, echo=FALSE, cache=FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print="75")
display_code <- FALSE
knitr::opts_knit$set(global.par = TRUE, width=75)
knitr::opts_chunk$set(
                echo    = display_code,
                prompt  = FALSE,
                tidy    = TRUE,
                comment = NA,
                message = FALSE,
                warning = FALSE)
                #cache=TRUE)

library(rmdformats)
library(runGOA)
library(magrittr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(beanplot)
palette(brewer.pal(6, "Set2"))

par(mgp = c(2, 0.5, 0))
par(mar = par()$mar * 0.7)
#load_all()
min_appearances <- 20
```

```{r DT_function}
dt_setup <-  function(data, n_rows = 10, lineHeight="80%") {
  DT::datatable(data,
                rownames = FALSE,
                options = list(dom = 't', 
                              pageLength = n_rows)
  ) %>%
  DT::formatStyle(0, target = 'row', lineHeight = lineHeight)
}

dark_b_blue <- "#374772" 
light_b_blue <- "#3cc1f2"
```

### Generate lists of human genes biased by the the chromosome that they're on.

Use a gene info file derived from a gtf file from Ensembl containing all the genes
and transcripts in the human genome.

<div class = "row">
<div class = "col-md-4">

```{r import_genfo_file}
# The gene info file contains the number of transcripts of all the genes in the human genome. 
# The processing of it is described in the processing_gtf file.
load("../data/genfo.rda")
load("../data/all_go_categories_human.rda")

all_GO_annotated_genes <- unique(as.vector(unlist(all_go_categories_human)))

genfo_all <- genfo

genfo <- genfo %>%
    filter(gene_name %in% all_GO_annotated_genes)

# count number of duplicate gene names - decide what to do about them...
genfo %>%
  count(gene_name) %>%
  filter(n > 1) %>%
  arrange(desc(n)) %>%
  dt_setup(n_rows = 8)
```

</div>
<div class = "col-md-1"></div>
<div class = "col-md-7">
<br><br>
This table show the number of times that a gene name appears in the genome annotated 
as a distinct gene. <br>
There are lots of duplicate gene names so we'll use the Ensembl gene IDs for the 
initial processing. <br> 
The gene set files that we use for the GO analysis do not contain Ensembl IDs so
we'll have to convert back to gene name at the point when we run the GO analysis.
<br><br>
<font size ="1">The highly duplicated gene names probably aren't in the GO 
annotations much but it'd be worth checking.</font>
<br><br>
</div>
</div>
```{r dup_Ensembl_IDs}
duplicated_Ensembl_IDs <- genfo %>%
                            count(gene_id) %>%
                            filter(n > 1) %>%
                            count() %>%
                            pull()
```

There are `r duplicated_Ensembl_IDs` duplicate Ensembl IDs so we'll use these for 
filtering and convert back to gene name at the end of this analysis. 
<br><br>

```{r}
chr_order <- c(1:22, "MT", "X", "Y")
```

## Number of genes per chromosome

<div class = "row">
<div class = "col-md-6">
```{r}
genfo <- genfo %>%
    select(gene_id, gene_name, chromosome) 

dt_setup(genfo)
```

</div>
<div class = "col-md-6">
```{r}
no_per_chr <- genfo %>%
    count(chromosome)# %>%
   # mutate(chromosome =  forcats::fct_reorder(chromosome, chr_order)) %>%
  #arrange(chromosome)
    #arrange(forcats::fct_reorder(chromosome, chr_order))

dt_setup(no_per_chr)
```
</div>
</div>

```{r helper_functions}
get_random_sample_for_category <- function(category, n_genes_to_select = 200) {
    
    filtered <- filter(genfo, chromosome == category)
    
    if(nrow(filtered) > n_genes_to_select){
    
      genes <- filtered  %>%
          sample_n(size = n_genes_to_select) %>%
          pull(gene_name)
    
    } else{  
      
      genes <- pull(filtered, gene_name)
    }  
  
    genes  
}       
```


```{r GO_analysis, eval = FALSE}
# This code is run once to get the GO analysis results- it takes a while to run
# it 100 times for each bias category.

# There are some duplicate gene names - MATR3 has 11 transcripts and 28 transcripts, TBCE has 19 and 57.
# Maybe this doesn't matter here, we're seeing what comes up when selecting genes with high transcripts. 

#load("../data/all_go_categories_human.rda")
load("../data/all_background_genes_human.rda")

# set up the summary files so we can populate them as we get the GO analysis results
go_category_counts <- tibble(GO_category = names(all_go_categories_human))

no_cats <- length(chr_order)

summary_stats <- tibble(
                      run_number         = rep(1:100, times = no_cats), 
                      bias_category      = rep(chr_order[1:no_cats], each = 100), 
                      sig_number_results = 0)

summary_stats$sig_number_results %<>% as.integer

# more efficient to pass this as an argument than having to calculate it each time
# the overrep test is run
unique_genes_in_categories <- unique(as.vector(unlist(all_go_categories_human)))

#for (i in 1:length(thresholds$category)){
#the very_very high category only has 183 genes so there's no point in randomly sampling
#that for 200 genes, 100 times. 
#
#
#!! This needs simplifying - there's too much in here to go wrong - latest was
#Error: Names must be unique. 
for (i in 1:no_cats){  
   
    chr <- i 
    bias_category <- paste0("chr", chr_order[i])
    go_category_counts <- mutate(go_category_counts, occurrences = 0)

    for(j in 1:100) {

        genes <- get_random_sample_for_category(chr)
        
        go_results <- overrep_test(all_go_categories_human, 
                                   query_genes = genes, 
                                   all_background_genes_human, 
                                   unique_genes_in_categories = unique_genes_in_categories,
                                   min_genes_in_category = 1, 
                                   max_genes_in_category = 500000)
        
        if(length(go_results) > 0){
           
            summary_stats <- summary_stats %>%
                mutate(sig_number_results = if_else(
                    bias_category == !!chr & run_number == j, 
                    nrow(go_results), 
                    sig_number_results
                ))
        }  
        go_category_counts <- go_category_counts %>%
            mutate(occurrences = if_else(GO_category %in% rownames(go_results), 
                                         occurrences + 1, 
                                         occurrences)
                   ) 
    }

    # I couldn't get tidyverse to work properly with mutating column names and adding to them, so we'll rename afterwards
    go_category_counts <- rename(go_category_counts, !!bias_category := occurrences)
}

human_chr_biased  <- go_category_counts
human_chr_summary <- summary_stats

usethis::use_data(human_chr_biased, overwrite = TRUE)
usethis::use_data(human_chr_summary, overwrite = TRUE)
```

<div class = "row">
<div class = "col-md-6">
```{r}
load("../data/human_chr_biased.rda")
load("../data/human_chr_summary.rda")

category_counts <- human_chr_summary %>%
    filter(sig_number_results > 0) %>%
    group_by(bias_category) %>%
    count(bias_category)

dt_setup(category_counts)
```
</div>
<div class = "col-md-6">
<br><br>
The table shows the number of functional enrichment analyses that returned 1 or 
more significantly enriched categories from the 100 that were run. 
</div> 
</div>
</div> 
<br>
<div class = "row">
<div class = "col-md-6">
```{r fig.height = 4, fig.width = 10}
no_categories <- human_chr_summary %>% 
  select(bias_category) %>% 
  unique() %>% 
  count() %>% 
  pull()


human_chr_summary %>%
    ggplot(aes(x = bias_category, y = sig_number_results)) +
    geom_violin(fill = light_b_blue, colour = dark_b_blue, lwd = 0.8) +
    xlab("") +
    ylab("number of categories") +
    theme_bw() +
    theme(axis.text.x = element_text(size = 12)) #+
   # geom_label(data = category_counts, colour = dark_b_blue, fill = "lightgrey", aes(label = paste0("n=",n), x = 1:no_categories, y = 250)) #+
 # geom_boxplot(width = 0.005)
```

```{r}
all_identified_categories <- human_chr_biased %>%
  filter_at(vars(-GO_category), any_vars(. > 0)) 

dt_setup(all_identified_categories)
```

In total, `r nrow(all_identified_categories)` functional categories appeared 
at least once.

## Filter Categories

We don't want categories that only appeared a handful of times so we'll use an 
arbitrary value of `r min_appearances` as a threshold.

```{r}
results <- human_chr_biased %>%
  tidyr::pivot_longer(starts_with("chr"), names_to = "chromosome", values_to = "occurrences") %>%
  filter(occurrences >= min_appearances)

dt_setup(results)
```


